---
title: "Primary EC - Number Prediction"
subtitle: "Biological Data Science Curse - CompBio Coimbra"
authors: c("V.Haberl","F.Moroff")
output:
  ioslides_presentation:
    widescreen: yes
    smaller: yes
    logo: www/FCTUC_V_FundoClaro.png
    css: www/io.css
# output: pdf_document
date: "2022-10-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(seqinr)
#library(bio3d)
library(plotly)
library(ggplot2)
library(dplyr)
library(ggfortify)
library(e1071)
library(caTools)
# ?rcdk
```

## Goal:{.build}

Train a Machine learning Algorithm to predict Enzyme Comission Number (EC -number ) from Protein Sequence.

 - Sequence > Structure > Function
 - Sequence > (ML)      > Function

## EC - Numbers
| Class                | Reaction catalyzed                                                                                                        | Typical reaction                                | Enzyme example                          |
|:-------------|:-------------------------------|:-------------|:-------------|
| EC 1 Oxidoreductases | Oxidation/reduction reactions; transfer of H and O atoms or electrons from one substance to another                       | AH + B → A + BH (reduced) A + O → AO (oxidized) | Dehydrogenase, oxidase                  |
| EC 2 Transferases    | Transfer of a functional groupf from one substance to another. The group may be methyl-, acyl-, amino- or phosphate group | AB + C → A + BC                                 | Transaminase, kinase                    |
| EC 3 Hydrolases      | Formation of two products from a substrate by hydrolysis                                                                  | AB + H2O → AOH + BH                             | Lipase, amylase, peptidase, phosphatase |
| EC 4 Lyases          | Non-hydrolytic addition or removal of groups from substrates. C-C, C-N, C-O or C-S bonds may be cleaved                   | RCOCOOH → RCOH + CO2 or [X-A+B-Y] → [A=B + X-Y] | Decarboxylase                           |
| EC 5 Isomerases      | Intramolecule rearrangement, i.e. isomerization changes within a single molecule                                          | ABC → BCA                                       | Isomerase, mutase                       |
| EC 6 Ligases         | Join together two molecules by synthesis of new C-O, C-S, C-N or C-C bonds with simultaneous breakdown of ATP             | X + Y + ATP → XY + ADP + Pi                     | Synthetase                              |
| EC 7 Translocases    | Catalyse the movement of ions or molecules across membranes or their separation within membranes                          |                                                 | Transporter                             |
(from <https://www.abbexa.com/enzyme-commission-number> *altered*)

## Challenges

-   SAR -paradoxon: SAR: \_Structure--Activity *Relationship* ... refers to the fact that it is not the case that all similar molecules have similar activities.

## TEST FOR GIT

## Input Data: {.build}

### Creation of the training dataset:

To work on the implementation of the algorithm we queried the Uniprot Database for reviewed human enzyme sequences with know EC numbers, using their REST-Api.

### Creation of larger dataset:

To create a larger training and validation dataset we used the training and validation data that had been used in the creation of _ECPred_ a Java application for ECN prediction. The benefits in using this dataset was that it is already curated for machine learning, is very vast. The drawback was that the repository only contains UniProt accession numbers and not the sequences of proteins. For this reason we had to download the sequences and build the dataset ourselfs afterwards.

To keep the scope of the project reasonlable we focused only on the prediction of the main enzyme identifier number EC{1,2,3,4,5,6}


  
## Recreating the ECPred Main- dataset 

### Copy Folder with accession nubers from github
```{bash, eval= FALSE}
git clone --depth 1 --filter=blob:none --sparse https://github.com/cansyl/ECPred
cd ECPred
git sparse-checkout set ECPred\ Datasets
cd ./ECPred\ Datasets
rm -r -f !\(EC_Main*\)  # only keep main EC number classification
```

## The training data:{.smaller}

```{bash}
cd ./ECPred/ECPred\ Datasets
tree
```

## Modeling Methods

We trained and compared the performance of four different ML approaches:

  - LDN
  - random forest 
  - knn 
  - svm
  
## Used Features {.build}

Naive approach:
  only Amino acid composition (AAC) as a baseline .
  
 _Feature engineering:_

we created additional features to the training data
those where obtained by using prebuilt functions from the R packages *seqinr* and *protr*.

  - number of residues
  - pysio-chemical classes
  - isoelectricity
  - CTD Descriptors - Composition
  - polar surface area 


## read in accession numbers


```{r}
Enzymes<-c("Hydrolases","Isomerases","Ligases","Oxidoreductases","Transferases")
## negatives
path ="./ECPred/ECPred Datasets/EC_MainClass_NegativeTrain/"
filenames<-paste0(Enzymes,"_Negative_Train.txt")
Hydrolases_neg_train<-scan(paste0(path,filenames[1]),character(),quote="")
Hydrolases_neg_train<-Hydrolases_neg_train[-c(1:3)]
```

```{r protr copm,eval=FALSE}
length(Hydrolases_neg_train)
library(protr)
a<-Sys.time()

ids <- Hydrolases_neg_train[1:100]
prots <- getUniProt(ids)
prots <- removeGaps(prots)
prots <- subset(prots,unlist(lapply(prots, protcheck)))
prots# remove problematic sequences
prots <- lapply(prots,extractAAC ) #calcualte AAC
print(Sys.time()-a)
144*2/60
```



